MSnbase2 benchmarking

### 2016-12-23 docs/articles/Figures/MSnbase-io-out.tex
Merge branch 'master' into devel
* master: (24 commits)
  fix typos from d342c70 in README.Rmd
  update installation ignore
  fix typo [ci skip]
  udpate docs with verbose is FALSE
  set default MSnbase verbose to FALSE
  update docs
  update docs
  update news with new news format
  Move NEWS to NEWS.md
  Migrate io and dev vignettes to BiocStyle's html_document2 style
  removes notes
  mv devel vignette to BiocStyle html2
  use BiocStyle's html_docuement2 for benchmarking vignette
  Amend show method for MSnExp
  Small change to show method for MSnExp
  udpate docs
  update docs with Seb's s

%\VignetteEngine{knitr}
%\VignetteIndexEntry{MSnbase IO capabilities}
%\VignetteKeywords{Mass Spectrometry, Proteomics, Infrastructure }
%\VignettePackage{MSnbase-io}

\documentclass[12pt,a4paper,english]{scrartcl}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}
\usepackage[auth-sc]{authblk}
\usepackage{setspace}
\onehalfspacing

\usetikzlibrary{shapes,arrows,shadows,fit}

% caption formatting
\setcapindent{0em}
\setkomafont{captionlabel}{\sffamily\bfseries}
\setkomafont{caption}{\sffamily}

\newcommand{\R}{\texttt{R} }
\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\mbox{\normalfont\textsf{#1}}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
%% colors
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}

\usepackage{geometry}
\geometry{verbose,
  tmargin = 2.5cm,
  bmargin = 2.5cm,
  lmargin = 3.0cm,
  rmargin = 3.0cm}

\usepackage{hyperref}
\usepackage{breakurl}
\hypersetup{%
  pdfusetitle,
  bookmarks = {true},
  bookmarksnumbered = {true},
  bookmarksopen = {true},
  bookmarksopenlevel = 2,
  unicode = {true},
  breaklinks = {false},
  hyperindex = {true},
  colorlinks = {true},
  linktocpage = {true},
  plainpages = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  pdfstartview = {Fit},
  pdfpagemode = {UseOutlines},
  pdfview = {XYZ null null null}
}

@@

\author{
  Laurent Gatto\thanks{\email{lg390@cam.ac.uk}}
}

\affil{
  Cambridge Center for Proteomics\\
  University of Cambridge
}

@@

\title{\Rpackage{MSnbase} input/output capabilities}

\maketitle

%% Abstract and keywords %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip 0.3in minus 0.1in
\hrule
\begin{abstract}
  This vignette describes \Rpackage{MSnbase}'s input and output 
  capabilities.
\end{abstract}
\textit{Keywords}: Mass Spectrometry (MS), proteomics, infrastructure, IO.
\vskip 0.1in minus 0.05in
\hrule
\vskip 0.2in minus 0.1in
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \tableofcontents

\vspace{10mm}

<<knitr, include=FALSE, cache=FALSE>>=
library("knitr")
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRolocdata"))
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', 
               par = TRUE,
               prompt = TRUE,
               comment = NA)
options(replace.assign = TRUE, 
        width = 70)
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{Foreword.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}

\Rpackage{MSnbase}'s aims are to facilitate the reproducible analysis of 
mass spectrometry data within the \R environment, from raw data import and processing, 
feature quantification, quantification and statistical analysis of the results \citep{Gatto2012}.
Data import functions for several formats are provided and intermediate or final 
results can also be saved or exported. 
These capabilities are presented below.

\section{Data input}

\paragraph{Raw data} Data stored in one of the published \texttt{XML}-based 
formats. i.e. \texttt{mzXML} \citep{Pedrioli2004}, \texttt{mzData} \citep{Orchard2007} 
or \texttt{mzML} \citep{Martens2010}, 
can be imported with the \Rfunction{readMSData} method, which makes use of 
the \Rpackage{mzR} package to create \Robject{MSnExp} objects. 
The files can be in profile or centroided mode. 
See \Rfunction{?readMSData} for details.

\paragraph{Peak lists} Peak lists in the \texttt{mgf} 
format\footnote{\url{http://www.matrixscience.com/help/data\_file\_help.html\#GEN}} 
can be imported using the \Rfunction{readMgfData}. 
In this case, the peak data has generally been pre-processed by other software.
See \Rfunction{?readMgfData} for details.

\paragraph{Quantitation data} Third party software can be used to generate 
quantitative data and exported as a spreadsheet (generally comma or tab separated format). 
This data as well as any additional meta-data can be imported with the 
\Rfunction{readMSnSet} function. See \Rfunction{?readMSnSet} for details.

\Rpackage{MSnbase} also supports the \texttt{mzTab} 
format\footnote{\url{http://code.google.com/p/mztab/}}, a light-weight, tab-delimited 
file format for proteomics data developed within the Proteomics Standards Initiative (PSI).
\texttt{mzTab} files can be read into \R with \Rfunction{readMzTabData} to create and \Robject{MSnSet} instance.

\begin{figure}[!htb] %% input
  \begin{center}
    \begin{tikzpicture}[node distance = 2cm, auto]
      % Place nodes
      \node [input] (raw) {Raw data in an open \texttt{XML} format};
      \node [fun, left of=raw, node distance=4cm] (readMSData) {\Rfunction{readMSData}};
      \node [input, below of=raw] (mgf) {Peak list in \texttt{mgf} format};
      \node [fun, left of=mgf, node distance=4cm] (readMgfData) {\Rfunction{readMgfData}};
      \node [input, below of=mgf] (spreadsheet) {Quantitation data as a spreadsheet};
      \node [input, below of=spreadsheet] (mztab) {\texttt{mzTab} format};
      \node [fun, left of=spreadsheet, node distance=4cm] (readMSnSet) {\Rfunction{readMSnSet}};
      \node [fun, left of=mztab, node distance=4cm] (readMzTabData) {\Rfunction{readMzTabData}};
      \node [obj, left of=readMSData, yshift=-2.5mm, node distance=4cm] (MSnExp) {\Robject{MSnExp}};
      \node [obj, left of=readMSnSet, node distance=4cm] (MSnSet) {\Robject{MSnSet}};
      \node [fun, below of=MSnExp] (quantify) {\Rfunction{quantify}};
      % Background
      \begin{pgfonlayer}{background}
        \node [fill=yellow!20,rounded corners, draw=black!50, dashed, fit=(MSnExp) (quantify) (MSnSet)] {};
      \end{pgfonlayer}
      % Draw edges
      \draw  (raw) -- (readMSData);
      \draw  (mgf) -- (readMgfData);
      \draw  (spreadsheet) -- (readMSnSet);
      \draw [->] (readMSData) -- (MSnExp);
      \draw [->] (readMgfData) -- (MSnExp);
      \draw [->] (readMSnSet) -- (MSnSet);
      \draw  (MSnExp) -- (quantify);
      \draw [->] (quantify) -- (MSnSet);
      \draw (mztab) -- (readMzTabData);
      \draw [->] (readMzTabData) -- (MSnSet);
    \end{tikzpicture}
    \caption{Illustration of \texttt{MSnbase} input capabilities. 
      The white and red boxes represent \R functions/methods and objects respectively. 
      The blue boxes represent different disk storage formats.}
    \label{fig:input}
  \end{center}
\end{figure}

\section{Data output}

\paragraph{RData files} \R objects can most easily be stored on disk with the \Rfunction{save} function.
It creates compressed binary images of the data representation that can later be read back from the file 
with the \Rfunction{load} function.

\paragraph{Peak lists} \Robject{MSnExp} instances as well as individual spectra can be written 
as \texttt{mgf} files with the \Rfunction{writeMgfData} method. Note that the meta-data in the original 
\R object can not be included in the file. See \Rfunction{?writeMgfData} for details.

\paragraph{Quantitation data} Quantitation data can be exported to spreadsheet files with 
the \Rfunction{write.exprs} method. Feature meta-data can be appended to the feature 
intensity values. See \Rfunction{?writeMgfData} for details.

\Robject{MSnSet} instances can also be exported to \texttt{mzTab} files using the 
\Rfunction{writeMzTabData} function.

\begin{figure}[!htb] %% out
  \begin{center}
    \begin{tikzpicture}[node distance = 2cm, auto]
      % Place nodes
      \node [input] (raw) {Raw data in an open \texttt{XML} format};
      \node [input, below of=raw] (mgf) {Peak list in \texttt{mgf} format};
      \node [fun, right of=mgf, node distance=4cm, yshift=1cm] (writeMgfData) {\Rfunction{writeMgfData}};
      \node [input, below of=mgf] (spreadsheet) {Quantitation data as a spreadsheet};
      \node [input, below of=spreadsheet] (mztab) {\texttt{mzTab} format};
      \node [fun, right of=spreadsheet, node distance=4cm] (writeexprs) {\Rfunction{write.exprs}};
      \node [fun, right of=mztab, node distance=4cm] (writeMzTabData) {\Rfunction{writeMzTabData}};
      \node [obj, right of=writeMgfData, yshift=1.25cm, node distance=4cm] (MSnExp) {\Robject{MSnExp}};
      \node [obj, right of=writeexprs, node distance=4cm] (MSnSet) {\Robject{MSnSet}};
      \node [fun, below of=MSnExp] (quantify) {\Rfunction{quantify}};
      % Background
      \begin{pgfonlayer}{background}
        \node [fill=yellow!20,rounded corners, draw=black!50, dashed, fit=(MSnExp) (quantify) (MSnSet)] {};
      \end{pgfonlayer}
      % Draw edges
      \draw  (MSnExp) -- (writeMgfData);
      \draw [->] (writeMgfData) -- (mgf);
      \draw (MSnSet) -- (writeexprs);
      \draw [->] (writeexprs) -- (spreadsheet);
      \draw  (MSnExp) -- (quantify);
      \draw [->] (quantify) -- (MSnSet);
      \draw (MSnSet) -- (writeMzTabData);
      \draw [->] (writeMzTabData) -- (mztab);
    \end{tikzpicture}
    \caption{Illustration of \texttt{MSnbase} output capabilities.
      The white and red boxes represent \R functions/methods and objects respectively. 
      The blue boxes represent different disk storage formats.}
    \label{fig:output}
  \end{center}
\end{figure}

\clearpage

\section{Creating \Robject{MSnSet} from text spread sheets}

This section describes the generation of \Robject{MSnSet} objects using 
data available in a text-based spreadsheet. This entry point into \R and 
\Rpackage{MSnbase} allows to import data processed by any of the 
third party mass-spectrometry processing software available and proceed 
with data exploration, normalisation and statistical analysis using 
functions available in \R and the numerous Bioconductor packages.

\subsection{A complete work flow}

The following section describes a work flow that uses three input files to 
create the \Robject{MSnSet}. These files respectively describe the quantitative 
expression data, the sample meta-data and the feature meta-data. 
It is taken from the \Rpackage{pRoloc} tutorial and uses example files from 
the \Rpackage{pRolocdata} package.

\subsubsection{The original data file}\label{sec:orgcsv}

We start by describing the  \texttt{csv} to be used as input using the \Rfunction{read.csv} function.

<<readCsvData0, tidy = FALSE>>=
## The original data for replicate 1, available
## from the pRolocdata package
f0 <- dir(system.file("extdata", package = "pRolocdata"), 
          full.names = TRUE, 
          pattern = "pr800866n_si_004-rep1.csv")
csv <- read.csv(f0)
@

The three first lines of the original spreadsheet, containing the data for replicate one, are illustrated below (using the function \Rfunction{head}). It contains \Sexpr{nrow(csv)} rows (proteins) and \Sexpr{ncol(csv)} columns, including protein identifiers, database accession numbers, gene symbols, reporter ion quantitation values, information related to protein identification, \ldots

<<showOrgCsv>>=
head(csv, n=3)
@

Below read in turn the spread sheets that contain the quantitation data (\texttt{exprsFile.csv}),
feature meta-data (\texttt{fdataFile.csv}) and sample meta-data (\texttt{pdataFile.csv}).

<<readCsvData1, tidy = FALSE>>=
## The quantitation data, from the original data
f1 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "exprsFile.csv")
exprsCsv <- read.csv(f1)
## Feature meta-data, from the original data
f2 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "fdataFile.csv")
fdataCsv <- read.csv(f2)
## Sample meta-data, a new file
f3 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "pdataFile.csv")
pdataCsv <- read.csv(f3)
@

\begin{description}

\item[\texttt{exprsFile.csv}] containing the quantitation (expression) data for the \Sexpr{nrow(exprsCsv)} proteins and 4 reporter tags.
<<showExprsFile>>=
head(exprsCsv, n=3)
@
\item[\texttt{fdataFile.csv}] containing meta-data for the \Sexpr{nrow(fdataCsv)} features (here proteins).
<<showFdFile>>=
head(fdataCsv, n=3)
@
\item[\texttt{pdataFile.csv}] containing samples (here fractions) meta-data. This simple file has been created manually.
<<showPdFile>>=
pdataCsv
@
\end{description}

The self-contained \Robject{MSnSet} can now easily be generated using the \Rfunction{readMSnSet} constructor, 
providing the respective \texttt{csv} file names shown above and specifying that the data is comma-separated 
(with \Robject{sep = ","}). Below, we call that object \Robject{res} and display its content.

<<makeMSnSet, tidy = FALSE>>=
library("MSnbase")
res <- readMSnSet(exprsFile = f1,
                  featureDataFile = f2,
                  phenoDataFile = f3,
                  sep = ",")
res
@

\subsubsection{The \Robject{MSnSet} class}

Although there are additional specific sub-containers for additional meta-data 
(for instance to make the object MIAPE compliant), the feature 
(the sub-container, or slot \Robject{featureData}) and sample (the \Robject{phenoData} slot) 
are the most important ones. They need to meet the following validity requirements 
(see figure \ref{fig:msnset}):

\begin{itemize}
\item the number of row in the expression/quantitation data and feature data must 
  be equal and the row names must match exactly, and  
\item the number of columns in the expression/quantitation data and number of row 
  in the sample meta-data must be equal and the column/row names must match exactly. 
\end{itemize}

A detailed description of the \Robject{MSnSet} class is available by typing \Rfunction{?MSnSet} in the \R console.

\begin{figure}[!hbt]
\centering
    \includegraphics[width=0.5\textwidth]{msnset.png}
\caption{Dimension requirements for the respective expression, feature and sample meta-data slots. }
\label{fig:msnset}
\end{figure}

The individual parts of this data object can be accessed with their respective accessor methods:

\begin{itemize}
\item the quantitation data can be retrieved with \Rfunction{exprs(res)}, 
\item the feature meta-data with \Rfunction{fData(res)} and 
\item the sample meta-data with \Rfunction{pData(res)}. 
\end{itemize}

\subsection{A shorter work flow}

The \Rfunction{readMSnSet2} function provides a simplified import workforce. 
It takes a single spreadsheet as input (default is \texttt{csv}) and extract the 
columns identified by \Robject{ecol} to create the expression data, while the 
others are used as feature meta-data. \Robject{ecol} can be a \Robject{character} with 
the respective column labels or a numeric with their indices. In the former case, 
it is important to make sure that the names match exactly. Special characters like 
\texttt{'-'} or \texttt{'('} will be transformed by \R into \texttt{'.'} when the 
\texttt{csv} file is read in.
Optionally, one can also specify a column to be used as feature names. 
Note that these must be unique to guarantee the final object validity.

<<readMSnSet2>>=
ecol <- paste("area", 114:117, sep = ".")
fname <- "Protein.ID"
eset <- readMSnSet2(f0, ecol, fname)
eset
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Session information}
\label{sec:sessionInfo}

<<label=sessioninfo,results='asis',echo=FALSE,cache=FALSE>>=
toLatex(sessionInfo())
@

\bibliographystyle{plainnat}
\bibliography{MSnbase}

>>>>

\documentclass{minimal}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows,shadows,fit,automata}

%%% <
\usepackage{verbatim}
\usepackage[active,tightpage]{preview}
\PreviewEnvironment{tikzpicture}
\setlength\PreviewBorder{5pt}%
%%% >

\begin{comment}
  :Title:  MSnbase Output
  :Author: Laurent Gatto
\end{comment}

@@

@@

\begin{tikzpicture}[node distance = 2cm, auto]
  % Place nodes
  \node [input] (raw) {Raw data in an open \texttt{XML} format};
  \node [input, below of=raw] (mgf) {Peak list in \texttt{mgf} format};
  \node [fun, right of=mgf, node distance=4cm, yshift=1cm] (writeMgfData) {\texttt{writeMgfData}};
  \node [input, below of=mgf] (spreadsheet) {Quantitation data as a spreadsheet};
  % \node [input, below of=spreadsheet] (mztab) {\texttt{mzTab} format};
  \node [fun, right of=spreadsheet, node distance=4cm] (writeexprs) {\texttt{write.exprs}};
  % \node [fun, right of=mztab, node distance=4cm] (writeMzTabData) {\texttt{writeMzTabData}};
  \node [obj, right of=writeMgfData, yshift=1.25cm, node distance=4cm] (MSnExp) {\texttt{MSnExp}};
  \node [obj, right of=writeexprs, node distance=4cm] (MSnSet) {\texttt{MSnSet}};
  \node [fun, below of=MSnExp] (quantify) {\texttt{quantify}};
  % Background
  \begin{pgfonlayer}{background}
    \node [fill=yellow!20,rounded corners, draw=black!50, dashed, fit=(MSnExp) (quantify) (MSnSet)] {};
  \end{pgfonlayer}
  % Draw edges
  \draw  (MSnExp) -- (writeMgfData);
  \draw [->] (writeMgfData) -- (mgf);
  \draw (MSnSet) -- (writeexprs);
  \draw [->] (writeexprs) -- (spreadsheet);
  \draw  (MSnExp) -- (quantify);
  \draw [->] (quantify) -- (MSnSet);
  % \draw (MSnSet) -- (writeMzTabData);
  % \draw [->] (writeMzTabData) -- (mztab);
\end{tikzpicture}

### 2016-12-23 docs/articles/NoteAboutSpeedAndMemory.tex
Merge branch 'master' into devel
* master: (24 commits)
  fix typos from d342c70 in README.Rmd
  update installation ignore
  fix typo [ci skip]
  udpate docs with verbose is FALSE
  set default MSnbase verbose to FALSE
  update docs
  update docs
  update news with new news format
  Move NEWS to NEWS.md
  Migrate io and dev vignettes to BiocStyle's html_document2 style
  removes notes
  mv devel vignette to BiocStyle html2
  use BiocStyle's html_docuement2 for benchmarking vignette
  Amend show method for MSnExp
  Small change to show method for MSnExp
  udpate docs
  update docs with Seb's s

\paragraph{Speed and memory requirements} Raw mass spectrometry file are generally several 
hundreds of MB large and most of this is used for binary raw spectrum data. As such, 
data containers can easily grow very large and thus require large amounts of RAM. 
This requirement may be tackled in the future by avoiding to load the raw data into memory
and using random access to the content of \texttt{mzXML}/\texttt{mzML} data files on demand. 
When focusing on reporter ion quantitation, a direct solution for this is to trim the 
spectra using the \Robject{trimMz} method to select the area of interest and thus 
substantially reduce the size of the \Robject{Spectrum} objects. This is illustrated in 
section \ref{sec:trim} on page \pageref{trimMz-example} of the \texttt{MSnbase-demo} vignette.

>>>>

\paragraph{Speed and memory requirements} Raw mass spectrometry file are generally several 
hundreds of MB large and most of this is used for binary raw spectrum data. As such, 
data containers can easily grow very large and thus require large amounts of RAM. 
This requirement is being tackled by avoiding to load the raw data into memory
and using on-disk random access to the content of \texttt{mzXML}/\texttt{mzML} data files on demand. 
When focusing on reporter ion quantitation, a direct solution for this is to trim the 
spectra using the \Robject{trimMz} method to select the area of interest and thus 
substantially reduce the size of the \Robject{Spectrum} objects. This is illustrated in 
section \ref{sec:trim} on page \pageref{trimMz-example} of the \texttt{MSnbase-demo} vignette.
